---
title: "Choropleth Lab"
subtitle: "Week 10 Lab"
format:
  html:
    toc: true
    number-sections: true
execute:
  echo: true
  message: false
  warning: false
---

## 3.1 Identify Suitable Fill Variables for Choropleth Maps

| Variable | Variable type | Suitable as fill color? Why or why not? | Would the result be a choropleth map? Why or why not? |
|---|---|---|---|
| GDP per capita | Quantitative, intensive | Yes. It is normalized per person, so countries can be compared by color. | Yes. A choropleth is appropriate because fill color shows an intensive country-level statistic. |
| Life expectancy | Quantitative, intensive | Yes. It is an average measure, so countries can be compared by color. | Yes. It is a choropleth because color represents an intensive country-level statistic. |
| Region | Categorical, nominal | Yes, but only with a qualitative palette because regions are categories. | No. It would be a categorical thematic map, not a choropleth, because the colors do not show statistical intensity. |
| Population | Quantitative, extensive | Not ideal. Total population depends on country size and can be misleading as polygon fill. | No, not strictly. Population density would be better for a choropleth because it normalizes population by area. |

## 3.2 Start Coding
```{r}
#| label: setup
#| message: false
library(tidyverse)
library(mapview)
library(patchwork)
library(readxl)
library(rmapshaper)
library(scales)
library(sf)
library(shadowtext)
```

## 3.3. Import Country Statistics
```{r}
#| label: import-stats
country_stats <-
    "API_NY.GDP.PCAP.PP.KD_DS2_en_excel_v2_417139.xls" |>
    read_xls(skip = 3) |>
    select(name = "Country Name", code = "Country Code", gdp_per_cap = "2024") |>
    left_join(
        "API_SP.DYN.LE00.IN_DS2_en_excel_v2_404463.xls" |>
            read_xls(skip = 3) |>
            select(code = "Country Code", life_exp = "2024"),
        join_by(code)
    ) |>
    left_join(
        "API_NY.GDP.PCAP.PP.KD_DS2_en_excel_v2_417139.xls" |>
            read_xls(sheet = "Metadata - Countries") |>
            select(code = "Country Code", region = "Region"),
        join_by(code)
    ) |>
    filter(!is.na(region)) |>
    select(name, code, gdp_per_cap, life_exp)

country_stats
```

## 3.3. Verifier Task
```{r}
#| label: verify-3.1
glimpse(country_stats)
nrow(country_stats)
country_stats |>
    summarize(
        n_economies      = n(),
        n_missing_gdp    = sum(is.na(gdp_per_cap)),
        n_missing_life   = sum(is.na(life_exp)),
        n_missing_either = sum(is.na(gdp_per_cap) | is.na(life_exp)),
        n_complete       = sum(!is.na(gdp_per_cap) & !is.na(life_exp))
    )
```

## 3.4. Import the Boundaries as sf Objects
```{r}
#| label: read-boundaries
#| cache: true
country_boundaries <-
    "WB_GAD_ADM0_complete.geojson" |>
    read_sf() |>
    select(code = ISO_A3, status = WB_STATUS)

country_boundaries
```

## 3.4. Verifier Task
```{r}
#| label: verifier-3.4
class(country_boundaries)
unique(st_geometry_type(country_boundaries))
st_crs(country_boundaries)$input
st_is_longlat(country_boundaries)
```

## 3.5 Validate and Repair the Geometries
```{r}
#| label: repair-geometries
#| cache: true
sum(!st_is_valid(country_boundaries))

country_boundaries <- st_make_valid(country_boundaries)

sum(!st_is_valid(country_boundaries))
```

## 3.5. Verifier Task
```{r}
#| label: verify-3.5
#| cache: true

# Before
n_rows_before <- nrow(country_boundaries)
geom_types_before <- unique(st_geometry_type(country_boundaries))
sum(!st_is_valid(country_boundaries))
country_boundaries <- st_make_valid(country_boundaries)

# After
sum(!st_is_valid(country_boundaries))
identical(n_rows_before, nrow(country_boundaries))
identical(geom_types_before, unique(st_geometry_type(country_boundaries)))
```

## 3.6. Simplify the Polygons
```{r}
#| label: simplify
# a. Number of boundary points before simplification
points_before <- npts(country_boundaries)
points_before

# b. Number of polygons for Singapore before simplification
sgp_polygons_before <- country_boundaries |>
  filter(code == "SGP") |>
  st_cast("POLYGON") |>
  nrow()

sgp_polygons_before

# c. Simplify to approximately 10,000 boundary points
keep_target <- 10000 / points_before

country_boundaries <- country_boundaries |>
  ms_simplify(
    keep = keep_target,
    keep_shapes = TRUE
  )

# d. Number of polygons for Singapore after simplification
points_after <- npts(country_boundaries)
points_after

sgp_polygons_after <- country_boundaries |>
  filter(code == "SGP") |>
  st_cast("POLYGON") |>
  nrow()

sgp_polygons_after
```

## 3.6. Verifier Task
```{r}
#| label: verify-3.6
nrow(country_boundaries)
sum(st_is_empty(country_boundaries))
sum(!st_is_valid(country_boundaries))
```

## 3.7. Cure the Antimeridian Problem
```{r}
#| label: antimeridian-prob
# a.
ggplot(country_boundaries) + geom_sf()

# b.
country_boundaries <-
    country_boundaries |>
    st_wrap_dateline() |>
    st_cast("MULTIPOLYGON")

ggplot(country_boundaries) + geom_sf()
```

## 3.7. Verifier Task
```{r}
#| label: verify-3.7
n_rows_before <- nrow(country_boundaries)
geom_types_before <- unique(st_geometry_type(country_boundaries))

country_boundaries <-
    country_boundaries |>
    st_wrap_dateline() |>
    st_cast("MULTIPOLYGON")

identical(n_rows_before, nrow(country_boundaries))
identical(geom_types_before, unique(st_geometry_type(country_boundaries)))
sum(!st_is_valid(country_boundaries))
```

## 3.8 Merge Features That Share a Country Code
```{r}
#| label: merge-country-code

# a. 
country_boundaries |>
    st_drop_geometry() |>
    filter(!is.na(code)) |>
    count(code) |>
    filter(n > 1)

# c. 
country_boundaries <-
    country_boundaries |>
    mutate(group_id = if_else(is.na(code), row_number(), NA_integer_)) |>
    group_by(code, group_id) |>
    summarize(
        status = if_else(
            any(status == "Member State"),
            "Contains a Member State",
            "Does Not Contain a Member State"
        ),
        .groups = "drop"
    ) |>
    select(-group_id) |>
    st_cast("MULTIPOLYGON")
```

b. The code BES is shared with Bonaire, Saba, and Sint Eustatius. These are separate Caribbean islands that have been special municipalities of the Netherlands since 2010 following the dissolution of the Netherlands Antilles. ^[https://www.iso.org/obp/ui/?__cf_chl_f_tk=ALBocD5LfG6qwIIA62NLdSfQ5zLB_W53YejqIETQukc-1783392780-1.0.1.1-MEBqGxLme6aD2gyXgD.eFst0MKrlMo1KMilegYKuQTU#iso:code:3166:BQ]
