## 3.1 Identify Suitable Fill Variables for Choropleth Maps

| Variable | Variable type | Suitable as fill color? Why or why not? | Would the result be a choropleth map? Why or why not? |
|---|---|---|---|
| GDP per capita | Quantitative, intensive | Yes. It is normalized per person, so countries can be compared by color. | Yes. A choropleth is appropriate because fill color shows an intensive country-level statistic. |
| Life expectancy | Quantitative, intensive | Yes. It is an average measure, so countries can be compared by color. | Yes. It is a choropleth because color represents an intensive country-level statistic. |
| Region | Categorical, nominal | Yes, but only with a qualitative palette because regions are categories. | No. It would be a categorical thematic map, not a choropleth, because the colors do not show statistical intensity. |
| Population | Quantitative, extensive | Not ideal. Total population depends on country size and can be misleading as polygon fill. | No, not strictly. Population density would be better for a choropleth because it normalizes population by area. |

## 3.2 Start Coding
```{r}
 #| label: setup
  #| message: false
  library(tidyverse)
  library(mapview)
  library(patchwork)
  library(readxl)
  library(rmapshaper)
  library(scales)
  library(sf)
  library(shadowtext)
```

```{r}
gdp_file <- "API_NY.GDP.PCAP.PP.KD_DS2_en_excel_v2_417139.xls"
life_file <- "API_SP.DYN.LE00.IN_DS2_en_excel_v2_404463.xls"

gdp <- read_xls(gdp_file, sheet = "Data", skip = 3) |>
transmute(
    name = `Country Name`,
    code = `Country Code`,
    gdp_per_cap = as.numeric(`2024`)
)

life <- read_xls(life_file, sheet = "Data", skip = 3) |>
transmute(
    name = `Country Name`,
    code = `Country Code`,
    life_exp = as.numeric(`2024`)
)

regions <- read_xls(gdp_file, sheet = "Metadata - Countries") |>
transmute(
    code = `Country Code`,
    region = Region
)

country_stats <- gdp |>
left_join(life, by = c("name", "code")) |>
left_join(regions, by = "code") |>
filter(!is.na(region)) |>
select(name, code, gdp_per_cap, life_exp)

country_stats
nrow(country_stats)
```

```{r}
country_boundaries <- st_read("WB_GAD_ADM0_complete.geojson", quiet = TRUE) |>
    transmute(
      code = ISO_A3,
      status = WB_STATUS
    )

  country_boundaries
  nrow(country_boundaries)
  unique(st_geometry_type(country_boundaries))
  st_is_longlat(country_boundaries)
```

```{r}
sum(!st_is_valid(country_boundaries))

country_boundaries <- st_make_valid(country_boundaries)

sum(!st_is_valid(country_boundaries))
nrow(country_boundaries)
unique(st_geometry_type(country_boundaries))
```

```{r}
# a. Number of boundary points before simplification
points_before <- npts(country_boundaries)
points_before

# b. Number of polygons for Singapore before simplification
sgp_polygons_before <- country_boundaries |>
  filter(code == "SGP") |>
  st_cast("POLYGON") |>
  nrow()

sgp_polygons_before

# c. Simplify to approximately 10,000 boundary points
keep_target <- 10000 / points_before

country_boundaries <- country_boundaries |>
  ms_simplify(
    keep = keep_target,
    keep_shapes = TRUE
  )

# d. Number of polygons for Singapore after simplification
points_after <- npts(country_boundaries)
points_after

sgp_polygons_after <- country_boundaries |>
  filter(code == "SGP") |>
  st_cast("POLYGON") |>
  nrow()

sgp_polygons_after

# Verifier checks
nrow(country_boundaries)
sum(st_is_empty(country_boundaries))
sum(!st_is_valid(country_boundaries))
```

