---
title: "Choropleth Lab"
subtitle: "Week 10 Lab"
format:
  html:
    toc: true
    number-sections: false
execute:
  echo: true
  message: false
  warning: false
---

## 3.1. Identify Suitable Fill Variables for Choropleth Maps

| Variable | Variable type | Suitable as fill color? Why or why not? | Would the result be a choropleth map? Why or why not? |
|---|---|---|---|
| GDP per capita | Quantitative, intensive | Yes. It is normalized per person, so countries can be compared by color. | Yes. A choropleth is appropriate because fill color shows an intensive country-level statistic. |
| Life expectancy | Quantitative, intensive | Yes. It is an average measure, so countries can be compared by color. | Yes. It is a choropleth because color represents an intensive country-level statistic. |
| Region | Categorical, nominal | Yes, but only with a qualitative palette because regions are categories. | No. It would be a categorical thematic map, not a choropleth, because the colors do not show statistical intensity. |
| Population | Quantitative, extensive | Not ideal. Total population depends on country size and can be misleading as polygon fill. | No, not strictly. Population density would be better for a choropleth because it normalizes population by area. |

## 3.2. Start Coding
```{r}
#| label: setup
#| message: false
library(tidyverse)
library(mapview)
library(patchwork)
library(readxl)
library(rmapshaper)
library(scales)
library(sf)
library(shadowtext)
```

## 3.3. Import Country Statistics
```{r}
#| label: import-stats
country_stats <-
  "API_NY.GDP.PCAP.PP.KD_DS2_en_excel_v2_417139.xls" |>
  read_xls(skip = 3) |>
  select(name = "Country Name", code = "Country Code", gdp_per_cap = "2024") |>
  left_join(
    "API_SP.DYN.LE00.IN_DS2_en_excel_v2_404463.xls" |>
      read_xls(skip = 3) |>
      select(code = "Country Code", life_exp = "2024"),
    join_by(code)
  ) |>
  left_join(
    "API_NY.GDP.PCAP.PP.KD_DS2_en_excel_v2_417139.xls" |>
      read_xls(sheet = "Metadata - Countries") |>
      select(code = "Country Code", region = "Region"),
    join_by(code)
  ) |>
  filter(!is.na(region)) |>
  select(name, code, gdp_per_cap, life_exp)

country_stats
```

## 3.3. Verifier Task
```{r}
#| label: verify-3.1
glimpse(country_stats)
nrow(country_stats)
country_stats |>
  summarize(
    n_economies = n(),
    n_missing_gdp = sum(is.na(gdp_per_cap)),
    n_missing_life = sum(is.na(life_exp)),
    n_missing_either = sum(is.na(gdp_per_cap) | is.na(life_exp)),
    n_complete = sum(!is.na(gdp_per_cap) & !is.na(life_exp))
  )
```

## 3.4. Import the Boundaries as sf Objects
```{r}
#| label: read-boundaries
#| cache: true
country_boundaries <-
  "WB_GAD_ADM0_complete.geojson" |>
  read_sf() |>
  select(code = ISO_A3, status = WB_STATUS)

country_boundaries
```

## 3.4. Verifier Task
```{r}
#| label: verifier-3.4
class(country_boundaries)
unique(st_geometry_type(country_boundaries))
st_crs(country_boundaries)$input
st_is_longlat(country_boundaries)
```

## 3.5. Validate and Repair the Geometries
```{r}
#| label: repair-geometries
#| cache: true
sum(!st_is_valid(country_boundaries))

country_boundaries <- st_make_valid(country_boundaries)

sum(!st_is_valid(country_boundaries))
```

## 3.5. Verifier Task
```{r}
#| label: verify-3.5
#| cache: true

# Before
n_rows_before <- nrow(country_boundaries)
geom_types_before <- unique(st_geometry_type(country_boundaries))
sum(!st_is_valid(country_boundaries))
country_boundaries <- st_make_valid(country_boundaries)

# After
sum(!st_is_valid(country_boundaries))
identical(n_rows_before, nrow(country_boundaries))
identical(geom_types_before, unique(st_geometry_type(country_boundaries)))
```

## 3.6. Simplify the Polygons
```{r}
#| label: simplify
# a. Number of boundary points before simplification
points_before <- npts(country_boundaries)
points_before

# b. Number of polygons for Singapore before simplification
sgp_polygons_before <- country_boundaries |>
  filter(code == "SGP") |>
  st_cast("POLYGON") |>
  nrow()

sgp_polygons_before

# c. Simplify to approximately 10,000 boundary points
keep_target <- 10000 / points_before

country_boundaries <- country_boundaries |>
  ms_simplify(
    keep = keep_target,
    keep_shapes = TRUE
  )

# d. Number of polygons for Singapore after simplification
points_after <- npts(country_boundaries)
points_after

sgp_polygons_after <- country_boundaries |>
  filter(code == "SGP") |>
  st_cast("POLYGON") |>
  nrow()

sgp_polygons_after
```

## 3.6. Verifier Task
```{r}
#| label: verify-3.6
nrow(country_boundaries)
sum(st_is_empty(country_boundaries))
sum(!st_is_valid(country_boundaries))
```

## 3.7. Cure the Antimeridian Problem
```{r}
#| label: antimeridian-prob
# a.
ggplot(country_boundaries) + geom_sf()

# b.
country_boundaries <-
  country_boundaries |>
  st_wrap_dateline() |>
  st_cast("MULTIPOLYGON")

ggplot(country_boundaries) + geom_sf()
```

## 3.7. Verifier Task
```{r}
#| label: verify-3.7
n_rows_before <- nrow(country_boundaries)
geom_types_before <- unique(st_geometry_type(country_boundaries))

country_boundaries <-
  country_boundaries |>
  st_wrap_dateline() |>
  st_cast("MULTIPOLYGON")

identical(n_rows_before, nrow(country_boundaries))
identical(geom_types_before, unique(st_geometry_type(country_boundaries)))
sum(!st_is_valid(country_boundaries))
```

## 3.8. Merge Features That Share a Country Code
```{r}
#| label: merge-country-code

# a.
country_boundaries |>
  st_drop_geometry() |>
  filter(!is.na(code)) |>
  count(code) |>
  filter(n > 1)

# c.
country_boundaries <-
  country_boundaries |>
  mutate(group_id = if_else(is.na(code), row_number(), NA_integer_)) |>
  group_by(code, group_id) |>
  summarize(
    status = if_else(
      any(status == "Member State"),
      "Contains a Member State",
      "Does Not Contain a Member State"
    ),
    .groups = "drop"
  ) |>
  select(-group_id) |>
  st_cast("MULTIPOLYGON")
```

b. The code BES is shared with Bonaire, Saba, and Sint Eustatius. ^[[https://www.iso.org/obp/ui/?__cf_chl_f_tk=ALBocD5LfG6qwIIA62NLdSfQ5zLB_W53YejqIETQukc-1783392780-1.0.1.1-MEBqGxLme6aD2gyXgD.eFst0MKrlMo1KMilegYKuQTU#iso:code:3166:BQ](https://www.iso.org/obp/ui/?__cf_chl_f_tk=ALBocD5LfG6qwIIA62NLdSfQ5zLB_W53YejqIETQukc-1783392780-1.0.1.1-MEBqGxLme6aD2gyXgD.eFst0MKrlMo1KMilegYKuQTU#iso:code:3166:BQ)] These are separate Caribbean islands that have been special municipalities of the Netherlands since 2010 following the dissolution of the Netherlands Antilles. ^[[https://en.wikipedia.org/wiki/Caribbean_Netherlands#:~:text=These%20are%20the%20islands,islands%20for%20short.](https://en.wikipedia.org/wiki/Caribbean_Netherlands#:~:text=These%20are%20the%20islands,islands%20for%20short.)]


## 3.9. Join the statistics to the boundaries
```{r}
#| label: join-stats-to-boundaries
countries <- country_boundaries |>
  left_join(country_stats, by = "code")

countries
```

## 3.9. Verifier Task
```{r}
#| label: verify-3.9
# Check for any code in country_stats that didn't find a match in country_boundaries
unmatched <- country_stats |>
  anti_join(country_boundaries, by = "code")

unmatched
```

```{r}
#| label: confirm-countries
class(countries)
nrow(countries)
names(countries)
```

## 3.10. Compare Map projections

a. The default CRS is EPSG:4326. It uses longitude and latitude coordinates. It is not really a projected CRS, but when drawn as a flat map it appears like an equirectangular.

b. Equal Earth projection (EPSG:8857)

```{r}
ggplot(country_boundaries) +
  geom_sf() +
  coord_sf(crs = 8857)
```

c. Web Mercator
```{r}
ggplot(country_boundaries) +
  geom_sf() +
  coord_sf(crs = 3857)
```

d. Comparison 

Equal Earth (EPSG::8857):
It is designed to preserve area, keeps all countries at their true relative sizes. This is useful for global choropleth map because countries are compared using color, so the map should not make some countries look much larger than they really are. There is a low shape distortion near the poles but the shapes stay recognizable. -The curved lines of longitude and straight lines of latitude show that the map is a projection of the round Earth onto a flat surface.

Web Mercator (EPSG:3857)
The core problem is the area distortion, especially near the poles. So the places far from the equator are wildly inflated and the size is larger than they actually are. This can be misleading in a choropleth map because viewers may give more attention to contries that appear larger on the map. It preserves local shape quite well, which is useful for online maps and navigation. The graticule lines like rectangular grid, but this can hide how much the map is distorting the real size of countries, tricks viewers into misjudging relative sizes.

The best projection for a global choropleth map is the Equal Earth projection. Preserving relative area is the most important property for a choropleth when comparing countries using color. 

## 3.11. Draft a First Choropleth Map
```{r}
ggplot(countries) +
  geom_sf(aes(fill = gdp_per_cap)) +
  coord_sf(crs = "EPSG:8857")
```

**Findings from Figure 3.11:**

1. **Visibility of missing data:** Yes, countries with missing GDP data are still visible filled with the default grey color.
2. **Undetermined-status areas:** They are currently indistinguishable from data-missing countries. If an undetermined-status area lacks data, it is filled with the same default grey, making it impossible to tell apart from a regular country that simply has missing data.
3. **Color spread across GDP distribution:** The default continuous color scale does not spread its colors well. Because GDP per capita is highly right-skewed (a few countries have very high GDP while most have lower GDP), the linear color gradient causes most countries to appear in a similar dark color at the low end of the scale, making it difficult to differentiate between them.

## 3.12. Refine the Map
```{r}
gray_countries <- countries |> 
  filter(is.na(gdp_per_cap) | is.na(code)) |>
  mutate(
    map_category = if_else(is.na(code), "Undetermined", "Missing"),
    fill_color = if_else(is.na(code), "grey50", "grey80")
  )

ggplot() +
  geom_sf(
    data = filter(countries, !is.na(code), !is.na(gdp_per_cap)),
    aes(fill = gdp_per_cap)
  ) +
  geom_sf(
    data = gray_countries,
    aes(color = map_category), 
    fill = gray_countries$fill_color
  ) +
  scale_fill_fermenter(
    palette = "Oranges",
    direction = 1,
    breaks = c(2000, 5000, 10000, 20000, 50000, 100000),
    labels = label_number(scale_cut = cut_short_scale())
  ) +
  scale_color_manual(
    name = NULL,
    values = c("Missing" = "grey30", "Undetermined" = "grey30"),
    guide = guide_legend(override.aes = list(fill = c("grey80", "grey50")))
  ) +
  coord_sf(crs = "EPSG:8857")
```

## 3.13. Build the Two-Panel Choropleth Map
```{r}
#| label: final-map
#| fig-width: 10
#| fig-height: 12
#| out-width: 100%

# 1. Create the ocean background and side borders
# Globe background (Light blue fill, no border)
lon_top <- seq(-180, 180, by = 1)
lat_top <- rep(84, length(lon_top))

lon_right <- rep(180, length(seq(83, -56, by = -1)))
lat_right <- seq(83, -56, by = -1)

lon_bottom <- seq(180, -180, by = -1)
lat_bottom <- rep(-56, length(lon_bottom))

lon_left <- rep(-180, length(seq(-55, 84, by = 1)))
lat_left <- seq(-55, 84, by = 1)

lon <- c(lon_top, lon_right, lon_bottom, lon_left)
lat <- c(lat_top, lat_right, lat_bottom, lat_left)

ocean_bg <- st_polygon(list(cbind(lon, lat))) |> st_sfc(crs = 4326)

# Left and right antimeridian edges (Thin black line)
left_edge <- st_linestring(cbind(rep(-180, 141), seq(-56, 84, by = 1))) |> st_sfc(crs = 4326)
right_edge <- st_linestring(cbind(rep(180, 141), seq(-56, 84, by = 1))) |> st_sfc(crs = 4326)
side_edges <- c(left_edge, right_edge)

# 2. Get top 40 largest countries for labels
largest_40 <- countries |>
  filter(!is.na(code)) |>
  mutate(area = st_area(geometry)) |>
  arrange(desc(area)) |>
  slice_head(n = 40) |>
  st_centroid() |>
  st_transform(8857)

largest_40_coords <- largest_40 |>
  mutate(
    x = st_coordinates(geometry)[, 1],
    y = st_coordinates(geometry)[, 2]
  ) |>
  st_drop_geometry()

# 3. Y-limits in projected space (EPSG:8857)
lims <- st_sfc(st_point(c(0, -56)), st_point(c(0, 84)), crs = 4326) |> 
  st_transform(8857) |> 
  st_coordinates()
y_limits <- lims[, "Y"]

create_map <- function(var_name, fill_palette, breaks, legend_title, label_func) {
  
  main_data <- countries |>
    filter(!is.na(code), !is.na(.data[[var_name]]))
    
  gray_data <- countries |> 
    filter(is.na(code) | is.na(.data[[var_name]])) |>
    mutate(
      map_category = if_else(is.na(code), "Undetermined", "Missing"),
      fill_color = if_else(is.na(code), "grey50", "grey80")
    )
    
  ggplot() +
    # Add our new ocean background and side edges BEFORE drawing the countries
    geom_sf(data = ocean_bg, fill = "#e0f3f8", color = NA) +
    geom_sf(data = side_edges, color = "black", linewidth = 0.3) +
    geom_sf(data = main_data, aes(fill = .data[[var_name]]), color = "grey30", linewidth = 0.15) +
    geom_sf(data = gray_data, aes(color = map_category), fill = gray_data$fill_color, linewidth = 0.15) +
    geom_shadowtext(
      data = largest_40_coords,
      aes(x = x, y = y, label = code),
      size = 2.5,
      color = "black",
      bg.color = "white",
      bg.r = 0.15
    ) +
    scale_fill_fermenter(
      palette = fill_palette,
      direction = 1,
      breaks = breaks,
      labels = label_func,
      guide = guide_colorsteps(
        title = legend_title,
        title.position = "top",
        title.hjust = 0.5,
        barwidth = 20,
        barheight = 0.7
      )
    ) +
    scale_color_manual(
      name = NULL,
      values = c("Missing" = "grey30", "Undetermined" = "grey30"),
      guide = "none" 
    ) +
    # Apply our calculated y-limits here
    coord_sf(crs = "EPSG:8857", ylim = y_limits, expand = FALSE) +
    theme_void() +
    theme(
      legend.position = "top",
      legend.title = element_text(face = "bold", size = 11),
      legend.text = element_text(size = 9),
      plot.margin = margin(5, 5, 5, 5)
    )
}

p_life <- create_map(
  "life_exp", 
  "Greens", 
  c(60, 65, 70, 75, 80), 
  "Life Expectancy (Years)",
  label_number()
)

p_gdp <- create_map(
  "gdp_per_cap", 
  "Oranges", 
  c(2000, 5000, 10000, 20000, 50000, 100000), 
  "GDP per Capita (Current USD)",
  label_number(scale_cut = cut_short_scale())
)

dummy_gray <- data.frame(
  map_category = c("Missing", "Undetermined"),
  fill_color = c("grey80", "grey50")
)

p_legend <- ggplot(dummy_gray) +
  geom_rect(aes(xmin = 0, xmax = 0, ymin = 0, ymax = 0, color = map_category), fill = dummy_gray$fill_color) +
  scale_color_manual(
    name = NULL,
    values = c("Missing" = "grey30", "Undetermined" = "grey30"),
    guide = guide_legend(override.aes = list(fill = c("grey80", "grey50")), nrow = 1)
  ) +
  theme_void() +
  theme(
    legend.position = "bottom",
    legend.text = element_text(size = 11),
    plot.margin = margin(t = 10, b = 0)
  )

final_plot <- p_life / p_gdp / p_legend +
  plot_layout(heights = c(1, 1, 0.05)) +
  plot_annotation(
    title = "Global Development in 2024",
    caption = "\"Undetermined\" refers to an area's legal status. \n Source: World bank (indicators, boundaries, and classification of legal status)",
    theme = theme(
      plot.title = element_text(hjust = 0.5, face = "bold", size = 18, margin = margin(b = 15)),
      plot.caption = element_text(hjust = 1, size = 10, color = "grey40")
    )
  )

final_plot
```

## 3.13. Verifier Task: Audit Checklist

*   **Layout:** The figure uses `patchwork` (`p_life / p_gdp`) to correctly stack life expectancy on top and GDP per capita on the bottom.
*   **Map projection:** Both maps use the Equal Earth equal-area projection (`coord_sf(crs = "EPSG:8857")`).
*   **Clipping:** The maps are correctly clipped vertically using manually calculated projected limits for 56°S and 84°N (`ylim = y_limits`).
*   **Color palettes:** `scale_fill_fermenter` is used with ColorBrewer palettes (`Greens` for Life Expectancy, `Oranges` for GDP).
*   **Gray fills:** Missing data is filled with `grey80` and undetermined areas with `grey50`. 
*   **Country borders:** Thin borders (`linewidth = 0.15`) are drawn for all countries to keep neighbors visually distinct.
*   **Legend scales:** Appropriate breaks are specified for both variables and numbers are formatted compactly using `label_number(scale_cut = cut_short_scale())`.
*   **Legend placement:** Fill legends are positioned horizontally, centered above their respective maps which acts as a map title. A dedicated dummy plot (`p_legend`) was used to display the two gray keys at the bottom of the figure.
*   **Country labels:** The 40 largest countries by `st_area()` are labelled at their centroids using `geom_shadowtext()`.
*   **Caption:** The World Bank credit and the required definition for "undetermined" regions are present in the bottom right caption.
*   **Background and frame:** A custom `ocean_bg` globe polygon fills the map extent with light blue `#e0f3f8`, while `side_edges` accurately draws the thin black antimeridian curves on the left and right.

## 3.14. Conclusions and Comparison

**Conclusions drawn from the maps:**
By comparing the two maps, a reader can immediately observe a strong correlation between wealth and health. The North Side of the world: (North America, Western Europe, Australia/New Zealand) and parts of East Asia have both the highest GDP per capita and the highest life expectancies. On the other hand, there is a distinct group of poverty and lower life expectancy concentrated primarily in Sub-Saharan Africa. The maps clearly demonstrate that GDP and Life Expectancy are heavily regionalized.

**Advantages of the Choropleth Maps:**

* The Choropleth Map instantly reveals spatial patterns, regional clustering, and neighbor effects that a scatter plot may not be able to show. 
* Every country with data is represented in its true geographical location. However, in the Preston-curve plot, many points representing smaller nations overlap over one another, making them hard to identify.
* Non-technical readers can immediately observe the global divide without needing to interpret graphs or trend lines.

**Disadvantages of the Choropleth Maps (vs. the Preston-curve plot):**

* Choropleth Maps inherently emphasize physical landmass over human population. Large countries like Russia, Canada, or Australia dominate the map due to their land size, while densely populated but geographically small countries are much harder to see. The Preston-curve plot solves this by sizing points by population.
* While the maps are great at quickly establishing some sort of relationship, the exact mathematical nature of the correlation is lost. The Preston-curve plot explicitly shows the relationship (GDP on life expectancy), which is not possible to deduce by simply glancing between two separate maps.

## 4.1. Reflect on AI Use
ChatGPT was used as an assistive AI tool.

One useful AI suggestion came in Section 3.6. Our draft simplification code was based on Tut10's ms_simplify() call, but it did not guard against small islands being dropped entirely once their polygons shrank below the simplification threshold. The AI pointed out the keep_shapes = TRUE argument, which keeps at least one polygon per feature. Without it, Singapore's row could have vanished from country_boundaries instead of just losing its smaller islets.

One AI response that was wrong came in Section 3.13, when building the ocean background polygon for the final map. The AI's first version only placed dense coordinate points along the top and bottom edges of the polygon, then connected the corners with straight lines. When rendered in the Equal Earth projection, this caused the ocean background to cut off on the left and right sides instead of bulging out to match the curved antimeridian. We caught this by visually inspecting the rendered map and noticing the background did not cover the full globe. We reported the specific visual symptom back to the AI, which diagnosed the cause (straight lines cannot follow a curved projected edge) and corrected it by adding dense points along all four edges of the polygon.

## 4.2. Member’s Role and Contributions
- Joshua Dui: Coder
- Benjamin: Navigator
- Mei Yuen: Verifier
- Dannon: Constructive Critic and Reporter
- Danish: AI Liaison